
```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

# Introduction

```{r, fig.alt="Title image: Choosing Genomics Tools Written by: Candace Savonen. Part of the ITN (ITCR training Network) and created through the Johns Hopkins Data Science Lab", out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1YwxXy2rnUgbx_7B7ENH9wpDX-j6JpJz6lGVzOkjo0qY/edit#slide=id.gd422c5de97_0_0")
```

## Target Audience  

The course is intended for students in the biomedical sciences and researchers who have been given data and don't know what to do with it or would like an overview of the different genomic data types that are out there.

_This course is written for individuals who:_   

- Have genomic data and don’t know what to do with it.
- Want a basic overview of genomic data types.
- Want to find resources for processing and interpreting genomics data.

```{r, fig.alt="For individuals who: Have genomic data and don’t know what to do with it. Want a basic overview of their genomic data type. Want to find resources for processing and interpreting genomics data", out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1YwxXy2rnUgbx_7B7ENH9wpDX-j6JpJz6lGVzOkjo0qY/edit#slide=id.g116525eff64_0_96")
```

## Topics covered:

```{r, fig.alt=" ", out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1YwxXy2rnUgbx_7B7ENH9wpDX-j6JpJz6lGVzOkjo0qY/edit#slide=id.g11db7c97851_0_143")
```

## Motivation

Cancer datasets are plentiful, complicated, and hold untold amounts of information regarding cancer biology. Cancer researchers are working to apply their expertise to the analysis of these vast amounts of data but training opportunities to properly equip them in these efforts can be sparse. This includes training in reproducible data analysis methods.

Often students and researchers need to utilize genomic data to reach the next steps of their research but may not have formal training in computational methods or the basics of the genomic data they are attempting to utilize.
AVOCADO: Put some more here after this is fleshed out.

## Curriculum  

```{r, fig.alt = " ", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1YwxXy2rnUgbx_7B7ENH9wpDX-j6JpJz6lGVzOkjo0qY/edit#slide=id.g11db7c97851_0_143")
```

**Goal of this course:**  
Equip learners with tutorials and resources so they can process and interpret their genomic data in a way that helps them meet their goals and handle the data properly

**What is not the goal**  
Teach learners about choosing parameters or about the ins and outs of every genomic tool they might be interested in. This course is meant to connect people to other resources that will help them with the specifics of their genomic data.

## How to use the course

This course is designed to be a jumping off point to more specific resources based on a genomic data type the learner has in mind (or currently on their computer). We encourage learners to follow links to resources we provide and feel free to jump around to chapters that are most useful for them.
