
```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

# DNA Methods

## Learning Objectives

```{r, fig.alt = "Learning objectives This chapter will demonstrate how to: Understand the goals and data collection for DNA sequence collection and variant identification. Compare and contrast the following methods: DNA/SNP microarrays, Whole Genome Sequencing, Whole Exome Sequencing, and Targeted Sequencing", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1YwxXy2rnUgbx_7B7ENH9wpDX-j6JpJz6lGVzOkjo0qY/edit#slide=id.g12890ae15d7_0_71")
```

## What are the goals of analyzing DNA sequences?

```{r, fig.alt = "", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1YwxXy2rnUgbx_7B7ENH9wpDX-j6JpJz6lGVzOkjo0qY/edit#slide=id.g12890ae15d7_0_71")
```

# DNA/SNP microarrays

## Where SNP array data comes from

## Microarray data **strengths**:

## Microarray data **limitations**:

## More about SNP microarrays

# Whole Genome Sequencing

## Where whole genome sequencing data comes from

## WGS data **strengths**:

## WGS **limitations**:

## More about WGS

# Whole Exome Sequencing

## Where whole exome sequencing data comes from

## WXS data **strengths**:

## WXS **limitations**:

## More about WGS

# Targeted Sequencing

## Where targeted sequencing data comes from

## Targeted sequencing data **strengths**:

## Targeted sequencing data **limitations**:

## More about targeted sequencing 
