
```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

# Microarray Data

## Learning Objectives

```{r, fig.alt = "This chapter will demonstrate how to: Understand the very general basics of microarray data collection and processing workflow. Understand the limitations and strengths of microarray data in general.", out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1YwxXy2rnUgbx_7B7ENH9wpDX-j6JpJz6lGVzOkjo0qY/edit#slide=id.g144bc8d6a68_0_12")
```

## Summary of microarrays

Microarrays have been in use since before high throughput sequencing methods became more affordable and widespread, but they still can be a effective and affordable tool for genomic assays. Depending on your goals, microarray may be a suitable choice for your genomic study.

## Pros:

- Microarrays are much more affordable than high throughput sequencing which can allow you to run more samples and have more statistical power.
- Microarrays take less time to process than most highthroughput sequencing methods. 
- Microarrays are generally less computationally intensive to process and you can get your results more quickly.

## Cons:

-

All microarrays work on hybridization to sets of oligonucleotides on a chip. However, the preparation of the samples, and the oligonucleotides' hybridization targets vary depending on the assay and goals.


## How do microarrays work?

```{r, fig.alt = "", out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1YwxXy2rnUgbx_7B7ENH9wpDX-j6JpJz6lGVzOkjo0qY/edit#slide=id.g14492c87338_0_51")
```




## What types of arrays are there?

### SNP arrays

```{r, fig.alt = "", out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1YwxXy2rnUgbx_7B7ENH9wpDX-j6JpJz6lGVzOkjo0qY/edit#slide=id.g13438a9a5b2_0_16")
```

### Gene expression arrays

```{r, fig.alt = "", out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1YwxXy2rnUgbx_7B7ENH9wpDX-j6JpJz6lGVzOkjo0qY/edit#slide=id.g142e3de7ce8_0_8")
```

### DNA methylation arrays

```{r, fig.alt = "", out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1YwxXy2rnUgbx_7B7ENH9wpDX-j6JpJz6lGVzOkjo0qY/edit#slide=id.g144bc8d6a68_0_1")
```
