```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

# Whole Genome Sequencing Methods

## Learning Objectives

```{r, fig.alt = "The learning objectives for this course are to: 1 Define the uses and applications of WGS 2 Describe the steps for generating WGS data 3 Understand the data analysis workflow for WGS", out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1YwxXy2rnUgbx_7B7ENH9wpDX-j6JpJz6lGVzOkjo0qY/edit#slide=id.g12890ae15d7_0_51")
```
The learning objectives for this course are to explain the use and application of WGS for genomics studies, outline the technical steps in generating WGS data, and detail the processing steps for analyzing and interpreting WGS data.

## WGS Overview

```{r, fig.alt = "Whole genome sequencing overview, Process of determining entirety of DNA sequence of organism’s genome at single time. Includes sequencing all chromosomal data and DNA from mitochondria. Used to identify functional variants associated with disease", out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1YwxXy2rnUgbx_7B7ENH9wpDX-j6JpJz6lGVzOkjo0qY/edit#slide=id.g138a6ce16b7_35_8")
```
At its core, whole genome sequencing is a techinque to throughly analyze the entire DNA sequence of any organism's genome. This includes sequencing all genes both coding and non-coding and all mitochondrial DNA. WGS is beneficial for identifying new and previously established variants related to disease and the regulatory elements of the genome including promoters, enhancers, and silencers.  


## Advantages and Disadvantages of WGS

```{r, fig.alt = "Advantages and Disadvantages of WGS: Most complete account of individual variation, Ability to study: Structural rearrangements, Copy number variations, Insertion-Deletions, SNPs, Sequencing repeats, Coding, non-coding, and mitochondrial genome coverage, allows for discovery - identify causative variants; Disadvantages include higher cost and more resources for storing and analyzing data", out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1YwxXy2rnUgbx_7B7ENH9wpDX-j6JpJz6lGVzOkjo0qY/edit#slide=id.g138a6ce16b7_35_13")
```
Alternatives to WGS include whole exome sequencing or sequencing the genes expressed in the genome and targeted gene sequencing. The main advantages of WGS include the ability to comprehensively analyze all regions of a genome, which has been massively beneficial for studying rare and complex diseases, and the ability to study structural rearrangements, gene copy number alterations, insertions and deletions, single nucleotide polymorphisms (SNPs), and sequencing repeats. Some disadvantages include higher sequencing costs and the necessity for more robust storage and analysis solutions to manage the much larger data output generated from WGS.

## Comparing WGS to other modalities

```{r, fig.alt = "Comparing WGS to other sequencing modalities. The most common DNA sequencing techniques are described. Whole genome sequencing coverages all genes and non-coding DNA. 3.2 billion bases are covered when applied to human samples. This the most expensive of the techniques. Depth of coverage required for 99.9% sensitivity is 30X. Whole exome sequencing coverage is the exome or expressed genes. Approximately 45 million bases are sequenced. This is a cost-effective technique. The depth of coverage required for 99.9% sensitivity is 100X. Targeted gene panel sequencing coverages 50-500 genes. 20,000 to 62 million bases are sequenced. This is the most cost-effective technique. Depth of coverage is >500X.", out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1YwxXy2rnUgbx_7B7ENH9wpDX-j6JpJz6lGVzOkjo0qY/edit#slide=id.g138a6ce16b7_35_18")
```
Compared to WES and targeted gene sequencing, WGS is the most expensive but requires the lowest depth of coverage to achieve 99.9% sensitivity. In other words, WGS requires sequencing each region of the genome (3.2 billion bases) 30 times in order to confidently be able to pick up all possible meaningful variants.

Alternatively, WES is a more cost effective way to study the genome, focusing just on the genes that are actively expressed. In this case, each gene must be sequenced 100x for sufficient sensitivity to pick up all meaningful coding variants.

In targeted gene sequencing, a panel of 50-500 genes of interest are selected. This technique is very applicable for studying a set of specific genes of interest at great depth to identify all varieties of mutations within those specific genes. These genes must be sequenced at much greater depth (>500x) to confidently identify all meaningful variants.


## WGS Considerations

```{r, fig.alt = "WGS Considerations , Genome type/size, Coverage requirements, Tissue source: fresh tissue, FFPE, blood, Library preparation protocol: PCR vs PCR-free", out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1YwxXy2rnUgbx_7B7ENH9wpDX-j6JpJz6lGVzOkjo0qY/edit#slide=id.g138a6ce16b7_35_28")
```
Some important considerations for WGS include what genome you are studying and the size of this genome. Included in this considerations is whether this genome has been sequenced before and you will have a "reference" genome to compare your data against or whether you will have to make a reference genome yourself. In addition, the depth of coverage for sequencing is an important consideration. The typical recommendation for WGS coverage is 30x, but this is on the lower side and man researchers find it does not provide sufficient coverage compared to 50x. Other important considerations include the tissue source and whether genetic alterations were introduced during processing (as fixation for FFPE often does) that will need to be accounted for during data analysis. In addition, the library preparation method of DNA amplification via PCR is very important as PCR can often introduce duplicates that interfere with interpreting whether a mutant gene is truly frequent or just over amplified during sequencing preparation.


## WGS Pipeline Overview

```{r, fig.alt = "WGS pipeline overview: Step 1: DNA extraction from sample, Step 2: library preparation, Step 3: Sequencing, Step 4: Analysis including data processing from Fastq, aligning reads to generate a BAM file, identifying variants to create a final VCF file", out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1YwxXy2rnUgbx_7B7ENH9wpDX-j6JpJz6lGVzOkjo0qY/edit#slide=id.g138a6ce16b7_35_33")
```
This slide outlines the necessary steps to key steps in performing WGS. First, DNA is extracted from a specific sample type (tissue, blood samples, cells, FFPE blocks, etc.). Either traditional (involving phenol and chloroform) or commercial kits can be used for this first step. Next, the DNA sequencing libraries are prepared. This involves fragmenting the DNA, adding sequencing adapters, and DNA amplification if the input DNA is not of sufficient quantity. After sequencing, data is analyzed by converting and aligning reads to generate a BAM file. Many analysis tools will use the BAM file to identify variants, which then generates a VCF file.


## WGS Data Pre-processing

```{r, fig.alt = "WGS data pre-processing pipeline overview: Raw data from sequencing is transformed into a Fastq file, reads are aligned and a Bam file is created, the data is sorted and merged, duplicates are identified, and the base quality score is recalibrated to create a final BAM file ", out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1YwxXy2rnUgbx_7B7ENH9wpDX-j6JpJz6lGVzOkjo0qY/edit#slide=id.g138a6ce16b7_35_38")
```
To expand on the last slide, raw sequencing reads are first transformed into a fastq file. Then the sequencing reads are aligned to a reference genome to create a BAM file. This data is sorted and merged, and PCR duplicates are identified. The confidence that each read was sequenced correctly is reflected in the base quality score. This score must be recalibrated at this step before variants are called. A final BAM file is thus created. This can be used for future analysis steps include variant or mutation identification, which is outlined on the following slide.


## Commonly Used Tools

```{r, fig.alt = "Tools commonly used in WGS data analysis", out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1YwxXy2rnUgbx_7B7ENH9wpDX-j6JpJz6lGVzOkjo0qY/edit#slide=id.g138a6ce16b7_35_43")
```
The following link provides the data analysis pipeline written by researchers in the NCI division of the NIH and provides a helpful overview of the typical steps necessary for [WGS analysis](https://docs.gdc.cancer.gov/Data/Bioinformatics_Pipelines/DNA_Seq_Variant_Calling_Pipeline/).

Here are many of the tools and resources used by researchers for analyzing WGS data.
First, these are common tools for data pre-processing:

- [Bedtools](https://bedtools.readthedocs.io/en/latest/) including the bamtofastq function, which is the first step in converting data off the sequencer to a useable format for downstream analysis
- [Samtools](http://www.htslib.org/) including tools for converting fastq to BAM files while mapping genes to the genome, duplicate read marking, and sorting reads
- [Picard2](https://broadinstitute.github.io/picard/) including tools to covert fastq to SAM files, filter files, create indices, mark read duplicates, sort files, and merge files
- [GATK](https://gatk.broadinstitute.org/hc/en-us) is a comprehensive set of tools from the Broad Institute for analyzing many types of sequencing data. For pre-processing, the print read function is very beneficial for writing the reads from a BAM or SAM file that pass specific criteria to a new file

These are the common tools used for somatic and germline variant identification. Note many researchers will often use a combination of these tools to narrow down only variants that are identified using a combination of these analysis algorithms:
- [Mutect2](https://gatk.broadinstitute.org/hc/en-us/articles/9570422171291-Mutect2) This is a beneficial variant calling tool with functions including using a "panel of normals" to better compare disease samples to normal and filtering functions for samples with orientation bias artifacts (FFPE samples) called F1R2
- [Varscan 2](https://varscan.sourceforge.net/) This is a helpful tool that utilizes a heuristic/statistic approach to variant calling. It is generally used and has capabilities for calling multi-sample variants. It can also be used for identifying copy number alterations in tumor-normal pairs
- [MuSE](https://bioinformatics.mdanderson.org/public-software/muse/) This is a beneficial mutation calling tool when you have both tumor and normal datasets. It uses a Markov Substitution model for Evolution, which models the evolution of the allelic composition of the tumor and normal tissue at each reference base. 
- [SvABA](https://github.com/walaj/svaba) This tool is especially useful for calling insertions and deletions (indels) because it utilizes a String Graph Assembler to assemble aberrantly aligned sequence reads that reflect indels or structural variants. Indels can be difficult to detect with standard alignment-based variant callers. 
- [Strelka](https://github.com/Illumina/strelka) This is a small variant caller designed by illumina. It is used for identifying germline variants in cohorts of samples and somatic variants in tumor/normal sample pairs. 
- [SomaticSniper](https://gmt.genome.wustl.edu/packages/somatic-sniper/) SomaticSniper can be used to identify SNPs in tumor/normal pairs. It calculates the probability that the tumor and normal genotypes are different and reports this probability as a somatic score. 
- [Pindel](https://www.sanger.ac.uk/tool/pindel/) Pindel is a tool that uses a pattern growth approach to detect breakpoints of large deletions, medium size insertsion/inversion, tandem duplications. 

Here are commonly used variant calling annotation tools: 
- [Annovar](https://annovar.openbioinformatics.org/en/latest/) This is a helpful tool for annotating, filtering, and combining the output data from the above tools. It can be used for gene-based, region-based, or filter-based annotations. 
- [GENCODE](https://www.gencodegenes.org/#) This tool can be used to identify and classify gene features in human and mouse genomes 
- [dbSNP]()
- [Ensembl]()

These are some of the commonly used copy number variation analysis tools: 
- [GATK]()
- [AscatNGS]()
- [TITAN CNA]()
- [gGNV]()
- [BIC-seq2]()

Here are the tools frequently used to visualize the output data from the above tools: 
- [IGV]()
- [Maftools]()
- [Prism]()




## Resources for WGS

Online tutorials:
[Galaxy tutorials](https://training.galaxyproject.org/training-material/topics/sequence-analysis/) 
[NCI resources](https://gdc.cancer.gov/about-data/gdc-data-processing/genomic-data-processing)
[Bioinformaticsdotca tutorial](https://bioinformaticsdotca.github.io/CAN_2021_module4_lab)

Papers comparing analysis tools: 
[Comparative analysis of WGS pipelines to minimize false negative findings](https://www.nature.com/articles/s41598-019-39108-2)
[Quality control and integration of genotypes from two calling pipelines for whole genome sequence data in the Alzheimer's disease sequencing project](https://pubmed.ncbi.nlm.nih.gov/29857119/) 
[Comprehensive fundamental somatic variant calling and quality management strategies for human cancer genomes](https://pubmed.ncbi.nlm.nih.gov/32510555/) 

