```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

# Whole Genome Sequencing Methods

## Learning Objectives

The learning objectives for this course are to: 
1 Define the uses and applications of WGS
2 Describe the steps for generating WGS data
3 Understand the data analysis workflow for WGS

```{r, fig.alt = "The learning objectives for this course are to: 
1 Define the uses and applications of WGS
2 Describe the steps for generating WGS data
3 Understand the data analysis workflow for WGS", out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1YwxXy2rnUgbx_7B7ENH9wpDX-j6JpJz6lGVzOkjo0qY/edit#slide=id.g12890ae15d7_0_66](https://docs.google.com/presentation/d/1YwxXy2rnUgbx_7B7ENH9wpDX-j6JpJz6lGVzOkjo0qY/edit#slide=id.g12890ae15d7_0_51")
```

## WGS Overview 
```{r, fig.alt = "Whole genome sequencing overview 
Process of determining entirety of DNA sequence of organism’s genome at single time. 
Includes sequencing all chromosomal data and DNA from mitochondria. 
Used to identify functional variants associated with disease", out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1YwxXy2rnUgbx_7B7ENH9wpDX-j6JpJz6lGVzOkjo0qY/edit#slide=id.g138a6ce16b7_35_8")
```

## Benefits of WGS 

```{r, fig.alt = "Benefits of WGS: 
Most complete account of individual variation 
Ability to study: 
Structural rearrangements
Copy number variations
Insertion-Deletions
SNPs
Sequencing repeats 
Coding, non-coding, and mitochondrial genome coverage 
WGS compared to WES: 
limited vs entire genome
targeted view of protein-coding regions
WGS allows for discovery - identify causative variants
", out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1YwxXy2rnUgbx_7B7ENH9wpDX-j6JpJz6lGVzOkjo0qY/edit#slide=id.g138a6ce16b7_35_8](https://docs.google.com/presentation/d/1YwxXy2rnUgbx_7B7ENH9wpDX-j6JpJz6lGVzOkjo0qY/edit#slide=id.g138a6ce16b7_35_13")
```
## Comparing WGS to other modalities 

```{r, fig.alt = "Comparing WGS to other sequencing modalities 
The most common DNA sequencing techniques are described. Whole genome sequencing coverages all genes and non-coding DNA. 3.2 billion bases are covered when applied to human samples. This the most expensive of the techniques. Depth of coverage required for 99.9% sensitivity is 30X. 
Whole exome sequencing coverage is the exome or expressed genes. Approximately 45 million bases are sequenced. This is a cost-effective technique. The depth of coverage required for 99.9% sensitivity is 100X. Targeted gene panel sequencing coverages 50-500 genes. 20,000 to 62 million bases are sequenced. This is the most cost-effective technique. Depth of coverage is >500X. ", out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1YwxXy2rnUgbx_7B7ENH9wpDX-j6JpJz6lGVzOkjo0qY/edit#slide=id.g138a6ce16b7_35_8](https://docs.google.com/presentation/d/1YwxXy2rnUgbx_7B7ENH9wpDX-j6JpJz6lGVzOkjo0qY/edit#slide=id.g138a6ce16b7_35_18")
```
## WGS Considerations 

## WGS Pipeline Overview 

## WGS Data Pre-processing 

## Commonly Used Tools 

## Resources for WGS 





