```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

# Whole Genome Sequencing Methods

## Learning Objectives

The learning objectives for this course are to: 
1 Define the uses and applications of WGS
2 Describe the steps for generating WGS data
3 Understand the data analysis workflow for WGS

```{r, fig.alt = "The learning objectives for this course are to: 
1 Define the uses and applications of WGS
2 Describe the steps for generating WGS data
3 Understand the data analysis workflow for WGS", out.width = "100%", echo = FALSE}
ottrpal::include_slide("[https://docs.google.com/presentation/d/1YwxXy2rnUgbx_7B7ENH9wpDX-j6JpJz6lGVzOkjo0qY/edit#slide=id.g12890ae15d7_0_66](https://docs.google.com/presentation/d/1YwxXy2rnUgbx_7B7ENH9wpDX-j6JpJz6lGVzOkjo0qY/edit#slide=id.g12890ae15d7_0_51)")
```

## WGS Overview 

Whole genome sequencing (WGS) is the process of determining the entire DNA sequencing of organism’s genome at single time. This includes sequencing all chromosomes and mitochondrial DNA.. 
The purpose is to identifiy functional variants associated with disease.
Two of the major applications for WGS are in research and personalized medicine. 
WGS is used by researchers to investigate functional variants and identify whether those changes are related to disease. 
In addition, clinicians can utilize WGS for making treatment decisions based on the patient's genetic profile. 

##Features and Benefits of WGS 
WGS 

