
```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

# Single-cell RNA-seq

## Learning Objectives

```{r, fig.alt = "This chapter will demonstrate how to: Understand the basics of single cell RNA-Seq data collection and processing workflow. Identify the next steps for your particular single cell RNA-seq data. Formulate questions to ask about your single cell RNA-seq data
", out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1YwxXy2rnUgbx_7B7ENH9wpDX-j6JpJz6lGVzOkjo0qY/edit#slide=id.g15bed4cad37_396_1")
```

## Where single-cell RNA-seq data comes from

```{r, fig.alt = "As opposed to bulk RNA-seq which can only tell us about tissue level and within patient variation, single-cell RNA-seq is able to tell us cell to cell variation in transcriptomics including intra-tumor heterogeneity", out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1YwxXy2rnUgbx_7B7ENH9wpDX-j6JpJz6lGVzOkjo0qY/edit#slide=id.g15bed4cad37_396_6")
```

As opposed to bulk RNA-seq which can only tell us about tissue level and within patient variation, single-cell RNA-seq is able to tell us cell to cell variation in transcriptomics including intra-tumor heterogeneity.

Single cell RNA-seq can give us cell level transciptional profiles. Whereas bulk RNA-seq masks cell to cell heterogeneity. If your research questions require cell-level transcriptional information, single-cell RNA-seq will on interest to you.

```{r, fig.alt = "Single cell RNA-seq can give us cell level transciptional profiles. Whereas bulk RNA-seq masks cell to cell heterogeneity.", out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1YwxXy2rnUgbx_7B7ENH9wpDX-j6JpJz6lGVzOkjo0qY/edit#slide=id.g15bed4cad37_396_11")
```

## Single-cell RNA-seq data types

There are broadly two categories of single-cell RNA-seq data methods we will discuss.

- **Full length RNA-seq**: Individual cells are physically separated and then sequenced.
- **Tag Based RNA-seq**: Individual cells are tagged with a barcode and their data is separated computationally.

Depending on your goals for your single cell RNA-seq analysis, you may want to choose one method over the other.

```{r, fig.alt = "Full length single cellRNA-seq Pros: Can be paired end sequencing which has less 3' bias. More complete coverage of transcripts which may be better for transcript discovery purposes. Cons: Is not very efficient (96 wells per plate). Takes longer to run days/weeks depending on the sample size. Expensive.", out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1YwxXy2rnUgbx_7B7ENH9wpDX-j6JpJz6lGVzOkjo0qY/edit#slide=id.g15bed4cad37_396_25")
```

```{r, fig.alt = "Tag based single cell RNA-seq. Pros: Can profile up to millions of cells. Takes less computing power. File storage requirements are smaller. Much less expensive. Cons: More intense 3' bias. Coverage is not as deep as full length single cell RNA-seq", out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1YwxXy2rnUgbx_7B7ENH9wpDX-j6JpJz6lGVzOkjo0qY/edit#slide=id.g15bed4cad37_396_25")
```

(Material borrowed from [@AlexsLemonade2022]).

### Unique Molecular identifiers

Often Tag based single cell RNA-seq methods will include not only a cell barcode for cell identification but will also have a unique molecular identfier (UMI) for original molecule identification. The idea behind the UMIs is it is a way to have insight into the original snapshot of the cell and potentially combat PCR amplification biases.

```{r, fig.alt = "Tag based single cell RNA-seq. Pros: Can profile up to millions of cells. Takes less computing power. File storage requirements are smaller. Much less expensive. Cons: More intense 3' bias. Coverage is not as deep as full length single cell RNA-seq", out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1YwxXy2rnUgbx_7B7ENH9wpDX-j6JpJz6lGVzOkjo0qY/edit#slide=id.g15bed4cad37_396_41")
```

- [AlevinQC](https://bioconductor.org/packages/release/bioc/html/alevinQC.html)

## Processing steps for single cell RNA-seq Data

- [Alevin]
- [10X Cell]
- [Seurat]
- [Single Cell Genome Viewer](https://github.com/KrasnitzLab/SCGV)

## More reading about single-cell RNA-seq data

- Hemberg Lab [scRNA-seq training course](https://scrnaseq-course.cog.sanger.ac.uk/website/index.html)
- [ASAP: Automated Single-cell Analysis Pipeline](https://asap.epfl.ch/) is a web server that allows you to process scRNA-seq data. ([Gardeux _et al._ 2017.](https://doi.org/10.1093/bioinformatics/btx337 ))
- Smith. [_Unique Molecular Identifiers – the problem, the solution and the proof_](https://cgatoxford.wordpress.com/2015/08/14/unique-molecular-identifiers-the-problem-the-solution-and-the-proof/) provides an excellent background on UMIs.
- [Baran-Gale _et al._ (2018)](https://doi.org/10.1093/bfgp/elx035)  
- [Lafzi _et al._ (2019)](https://doi.org/10.1038/s41596-018-0073-y)
- [Ziegenhain _et al._ (2018)](http://dx.doi.org/10.1016/j.molcel.2017.01.023)
- [Zhang _et al._ (2018)](https://doi.org/10.1016/j.molcel.2018.10.020)
- [AlJanahi _et al._ (2018)](https://doi.org/10.1016/j.omtm.2018.07.003)
- [Amezquita _et al._ (2019)](https://www.nature.com/articles/s41592-019-0654-x)
- [Angerer _et al._ (2017)](http://dx.doi.org/10.1016/j.coisb.2017.07.004)
- [Bruning _et al._ (2021)](https://www.biorxiv.org/content/10.1101/2021.02.15.430948v2)
- [Luecken & Theis (2019)](https://doi.org/10.15252/msb.20188746)
