
```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

# ATAC-Seq

## Learning Objectives

```{r, fig.alt = "Learning objectives This chapter will demonstrate how to: Understand the basics of ATAC-Seq data collection and processing workflow. Identify the next steps for your particular ATAC-Seq data. Formulate questions to ask about your ATAC-Seq data", out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1YwxXy2rnUgbx_7B7ENH9wpDX-j6JpJz6lGVzOkjo0qY/edit#slide=id.g12890ae15d7_0_66")
```

## What are the goals of ATAC-Seq analysis?

The goals of ATAC-seq are to identify the accessible regions of the genome in a particular set of samples. These data allow us to understand the relationships between the chromatin accessibility patterns and cell states, and to understand the mechanistic causes and consequences of these chromatin accessibility patterns.  

ATAC-seq data is generated by fragmenting the genome with the Tn5 endonuclease and sequencing the shorter DNA fragments. While most of the genome is associated with protein complexes that preclude the digestion of DNA by Tn5, some regions of the genome have accessible chromatin that can be cleaved by Tn5 resulting in short (<500bp) fragments. These regions of the genome are of biological interest as they are likely to harbor transcription factor binding sites and to constitute cis-regulatory elements, genomic regions that are involved in the regulation of gene expression.


```{r, fig.alt = "What does accessibility to chromatin represent? In ATAC-seq we are able to sequence open chromatin and find out DNA sequences where chromatin is accessible for activity. ", out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1YwxXy2rnUgbx_7B7ENH9wpDX-j6JpJz6lGVzOkjo0qY/edit#slide=id.g14492c87338_0_23")
```

## ATAC-Seq general workflow overview

A basic ATAC-seq workflow involves mapping sequence reads to the genome, identifying peaks, assessing data quality, and identifying patterns of interest through clustering or identification of differentially accessible regions or other statistical means. Transcription factor binding motif enrichment or transcription factor ChIP-seq peaks enriched accessible chromatin regions.

###Data quality metrics:

#### Number of mapped reads
As for all DNA-sequencing based genomics technologies, a sufficient number of mapped reads is required to obtain meaningful results from a sample. You can read more about [general sequencing technologies in our previous chapter here](http://hutchdatascience.org/Choosing_Genomics_Tools/sequencing-data.html). For experiments on human samples this number should be greater than 20 million mapped unique reads.

#### Fragment size distribution:
ATAC-seq data is often generated using paired end sequencing technologies, which allow for characterization of ATAC-seq fragments. Histograms of these distributions using single base pair resolution bins reveal patterns of enrichment relative to the nucleosome scale of 147bp and the DNA-helix scale ~10.5bp.

#### Number of peaks:
Although the number of accessible chromatin regions can vary from one cell type to another, there are several regions that appear to be constitutively accessible across a most cell types. At least 20,000 peaks can be identified in a high quality experiment.

#### FRiP score (fraction of reads in peaks)
In high quality ATAC-seq data a large fraction of reads overlap with peaks, while in low quality data there is a high level of fragments that map to background regions.

#### Overlap with other chromatin accessibility data
Thousands of ATAC-seq samples have been produced in human and mouse. High quality ATAC-seq data will share a substantial proportion of peaks with many of these datasets.

#### Overlap with promoters
The promoter regions of many genes are constitutively accessible. Examining peak overlap with regions close to known protein coding gene transcription start sites can be used as a check for data quality.

## ATAC-Seq data **strengths**:

- ATAC-seq is broadly used by many laboratories to characterize accessible chromatin in cell lines or sorted cells derived from tissues.
- In principle, ATAC-seq can identify a large proportion of cis-regulatory elements.
- In contrast to ChIP-seq, ATAC-seq does not require specific antibodies, or knowledge of the important transcription factors in a system.

## ATAC-Seq data **limitations**:

- ATAC-seq does not precisely identify the  transcription factors or other chromatin associated factors that bind in or around chromatin accessible regions. This type of information needs to be inferred through analysis of transcription factor binding motif analysis or ChIP-seq data.
- Accessible regions are not necessarily cis-regulatory regions, although many of them are.
- The genes that are regulated by cis-regulatory elements cannot be identified conclusively by ATAC-seq alone.
- ATAC-seq data can be biased making direct comparisons difficult.
- ATAC-seq can be affected by batch effects.

## ATAC-Seq data considerations

The nucleosome is the fundamental unit of chromatin packaging in the genome and nucleosomal DNA is far less likely to be cleaved by the Tn5 nuclease than linker DNA. When DNA is fragmented by Tn5 the positions of the endpoints relative to the nucleosomes is an important consideration. When the ends are less than 147bp apart it is likely that both ends originate from the same linker region. Longer fragments can result from cuts on opposite sides of the same nucleosome, or even opposite sides of a genomic interval that encompasses multiple nucleosomes. The short fragments are therefore most likely to be nucleosome free and provide stronger evidence for transcription factor binding sites.

As will other genomics protocols, ATAC-seq data is subject to biases introduced in the ATAC-seq protocol and in the sequencing itself. Comparison of ATAC-seq data generated in different batches, by different laboratories or using different protocols might not be directly comparable. In addition, the Tn5 endonuclease does have biases in the precise DNA sequences it can cut. This should be taken into consideration when carrying out base pair resolution analyses including footprinting analysis and analysis of the effects of sequence variants on chromatin accessibility.   

Read depth will impact ATAC-seq signal, but enzyme strength and conditions can also alter the distribution of cuts.

When using ATAC-seq data to answer biological questions it is important to understand what types of bias could impact the results. To ensure valid results the analysis needs to use appropriate statistical methods, ensure enough high quality ATAC-seq data is available, including controls, and possibly reframing the questions.

## ATAC-seq analysis tools

- [MACS]()
- [CHIPS]()
- [Cistrome DB]()

## More resources about ATAC-seq data

- [ATAC-seq review](https://genomebiology.biomedcentral.com/articles/10.1186/s13059-020-1929-3)
- [Identifying and mitigating bias in chromatin](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4473780/)
- [CHIP Snakemake pipeline for analyzing ChIP-seq and chromatin accessibility data](https://f1000research.com/articles/10-517)
- [Paper on bias in DNase-seq footprinting analysis and fragment size effects, similar comments apply to ATAC-seq](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4018771/)
- [SELMA Method for evaluating footprint bias in ATAC-seq](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9492688/)
